The hidden manifold distance for functional data

### 2019-07-30 index.Rmd
more writing of methodology section

---
title: "The hidden manifold distance for functional data"
output:
  html_document:
    df_print: paged
    number_sections: yes
  word_document: default
  pdf_document:
    keep_tex: yes
    number_sections: yes
bibliography: references.bib
---

@@

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that the dimensionality of curves is only artifically high if they are smooth.
There is also another sense in which the dimensionality of the curves under study may be low. This is the much less explored idea that curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include classes of probability density functions and classes of warped curves of a common template function. 
Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration. 
In this work we set ourselves the task of estimation the pairwise geodesic distance between functional manifold data when we only have access to their noisy realizations which live near a non-linear manifold rather than exactly on the manifold. 
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
<!-- We will show that estimation of the geodesic distance cannot easily be accomplished by a naive method such as smoothing followed by a shortest-path algorithm. -->
The proposed methodology first projects the functional data on to the underlying hidden manifold and then performs operations on this hidden manifold. 
Good estimation of the pairwise geodesic distance has beneficial implications for many downstream tasks in functional data analysis which we illustrate in the case of distanced-based functional classification.

# Introduction 
<!-- Susan  -->
Many statistical methods depend on a measure of distance. Methods in functional data analysis, the study of infinite-dimensional smooth curves, is no exception. In situations where the manifold hypothesis might be plausible, one might consider using a geodesic distance that takes into account the intrinsic structure of the manifold rather than the standard $L^2$ distance. The benefits of adopting the geodesic distance may be realized in downstream tasks such as distanced-based clustering and classification of functional obvervations.

@@

In this work, we put forth a technique for the specific task of recovering pairwise geodesic distances $\{ d_\M (X_i,X_j)\}_{i>j}$ when we have, in lieu of $X_i$ and $X_j$, access only to discretely-observed noisy functional observations $Y_i$ and $Y_j$ that possibly live off the true manifold. This is accomplished by first projecting the functional data on to the underlying hidden manifold and then performing operations on this hidden manifold.

@@

## Related Work
<!-- Susan  -->
[@ChenMuller2012]

[@LinYao2017]

[@Dimeglio2014]

# Proposed method for estimating geodesic distances [Marie]

Suppose that each curve $X_i$ is observed with measurements errors on a grid $T_i=(t_{i1},\ldots,t_{iK})$, i.e. we observe a sample of $K$-dimensional vectors $Y_1,\ldots,Y_n$ with $Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}$, where the random variables $\epsilon_{ij}$ are of mean zero and uncorrelated with each other. We assume that the grids $T_1,\ldots,T_n$ are dense.

Let $\tilde X_1,\ldots,\tilde X_n$ denote the functional versions of the raw data by some smoothing method. Our proposed methodology is unspecific to the smoothing method employed. For simplicity however, let's say we employ spline smoothing to recover the functional versions of the raw data, i.e.\
$$ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}$$
where $\lambda>0$ is a tuning parameter controlling the smoothness of $\tilde X_i$.

We first describe an important helper algorithm which we call IsoGeo. It is a subroutine of the classic Isomap procedure.
Isomap is a three steps procedure that takes as input a set of points $x_1,\ldots,x_n\in \R^D$ and produces an embedding of the input data in the space $\R^d$ with $d<D$, that preserves pairwise geodesic distances.

\begin{enumerate}
\item The first step consists in constructing a weighted graph $G$ with nodes corresponding to the input data and for which the weight of an edge between two nodes $x_i$ and $x_j$ is equal to $||x_i-x_j ||_{\R^D}$. The graph $G$ is constructed such that a node is connected to a fix number $N$ of neighbors or to all nodes that are at a distance smaller than a given value $\epsilon$.
\item In the second step, the pairwise geodesic distances $d_\M(x_i,x_j)$
are estimated based on $G$. The geodesic distance between two nodes is simply the length of the shortest path in $G$ between these two nodes, which we calculate with the Floyd-Warshall algorithm [@Floyd1962].
<!-- Which one do we actually use in our implementation, Floyd-Warshall or Dijkstra's? -->
\item The last step consists in using MDS on the matrix of pairwise geodesic distances obtained in the first step in order to obtain an embedding in $\R^d$.
\end{enumerate}

In what follows, we call IsoGeo the procedure of estimating pairwise geodesic distances with the two first steps of Isomap. Note that Floyd or Dijkstra are methods to find the smallest path given a weigthed graph but they are not related to how to calculate that graph (which is the difficult part, depend on number of neighbors, many method out there to calculate that graph in a robust way, P-ISOMAP is one of these).

Our method is described by the following steps:

<!-- 1. [Susan: I think the smoothing step should be independent of our proposed methodology] Transform each vector $Y_i$ into a function $\tilde X_i$ by spline smoothing: -->
<!-- $$ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}$$ -->
<!-- where $\lambda>0$ is a tuning parameter controlling the smoothness of $\tilde X_i$.  -->
2. Obtain a $s$-dimensional representation $\tilde X^s_1,\ldots,\tilde X^s_n$ of the functions $\tilde X_1,\ldots,\tilde X_n$ that "preserves" the pairwise $L^2$ distances by using MDS.
3. Obtain $\tilde X^{s,\hat \M}_1,\ldots,\tilde X^{s,\hat \M}_n$ a "projection" of $\tilde X^s_1,\ldots,\tilde X^s_n$ onto a ridge $\hat \M$ which is computed with the subspace constrained mean-shift algorithm [@Ozertem2011].
4. Use  IsoGeo to approximate the pairwise geodesic distances $\{d_{\hat \M}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j)\}_{i>j}$ and define the $n$ by $n$ matrix 
$$
\hat G(i,j)=\hat G(j,i) = \left\{ \begin{array}{ll}
 d_{\hat \M}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j) & \textrm{if $i\neq j$,}\\
 0 & \textrm{otherwise.}
  \end{array} \right.
$$

Since the ridge estimation obtained from noisy measurements of a manifold should approximate well the manifold [@Genovese2014], we expect the points $\tilde X^{s,\hat \M}_1,\ldots,\tilde X^{s,\hat \M}_n$ to lie close to the real manifold $\M$ and then $d_{\hat \M}$ to be close to $d_\M$. Ridge estimation suffers from the curse of dimensionality, this is why we first reduce the dimension of our data with MDS and then apply the shift-mean algorithm to estimate the ridge.

## Selection of tuning parameters [Marie]
Describe heuristics. We pick the bandwidth $h$ in the subspace constrained mean-shift using Equation (A1) of [@ChenHo2015].

@@

* Near-$\epsilon$ isometry: area under the curve of the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.  
* Relative Frobenius: let $d$ and $\hat d$ be the true and estimated $n \times n$ geodesic distance matrices, respectively. 
$$
||d-\hat d||/||d||
$$
* Pearson correlation: take the upper diagonal of $d$ and $\hat d$, calculate their Pearson correlation coefficient.

>>>>

---
title: "The hidden manifold distance for functional data"
output:
  pdf_document:
    keep_tex: yes
    number_sections: yes
  word_document: default
  html_document:
    df_print: paged
    number_sections: yes
bibliography: references.bib
---

@@

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that the dimensionality of curves is only artifically high if they are smooth.
There is also another sense in which the dimensionality of the curves under study may be low. 
This is the much less explored idea that the curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include classes of probability density functions and classes of warped curves of a common template function. 
<!-- Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration.-->
In this work we address the estimation of the pairwise geodesic distance between functional manifold data when we only have access to their noisy realizations which live near, rather than exactly on, the manifold.
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
<!-- We will show that estimation of the geodesic distance cannot easily be accomplished by a naive method such as smoothing followed by a shortest-path algorithm. -->
The proposed methodology first projects the functional data on to the underlying hidden manifold and then performs operations on this hidden manifold. 
Good estimation of the pairwise geodesic distance has beneficial implications for many downstream tasks in functional data analysis which we illustrate in the case of distanced-based functional classification.

# Introduction 
Many statistical methods depend on a measure of distance. Methods in functional data analysis, the study of infinite-dimensional smooth curves, are no exception. Certain classes of functional data may naturally live on manifold, e.g.\ classes of probability density functions and classes of warped curves of a common template function.
In situations where the manifold hypothesis might be plausible, one might consider using a geodesic distance that takes into account the intrinsic structure of the manifold rather than the standard $L^2$ distance. The benefits of adopting the geodesic distance may be realized in downstream tasks such as distanced-based clustering and classification of functional obvervations.

@@

In this work, we put forth a technique for the specific task of recovering the $n{\times}n$ matrix $G$ of pairwise geodesic distances, i.e.\
$$
G(i,j)=G(j,i) = \left\{ \begin{array}{ll}
 \{ d_g (X_i,X_j)\}_{i>j} & \textrm{if $i\neq j$,}\\
 0 & \textrm{otherwise.}
  \end{array} \} \right.
$$ 
when we have, in lieu of $X_i$ and $X_j$, access only to discretely-observed noisy functional observations $Y_i$ and $Y_j$ that possibly live off the true manifold. This is accomplished by first projecting the functional data on to the underlying hidden manifold and then performing operations on this hidden manifold.

@@

## Related Work
The work of [@ChenMuller2012] was among the first in functional data analysis to consider the manifold hypothesis for functional data. A notion of the mean and variation of functional manifold data was introduced. Of particular relevance to our work is the modified ISOMAP procedure, called P-ISOMAP, introduced in [@ChenMuller2012], which adds a data-adaptive penalty to allow for noisy functional observations whereas in the classic ISOMAP algorithm, observations are assumed to lie exactly on the manifold. Another existing work for estimating the pairwise geodesic distance for functional manifold data can be found in [@Dimeglio2014]. The estimator is called Robust-ISOMAP for the fact that it is motivated by modifying the ISOMAP algorithm to be less sensitive to outliers. We will discuss both P-ISOMAP and robust ISOMAP in more details in the simulation section.

<!-- [@LinYao2017]  -->

# Proposed method for estimating geodesic distances

Suppose that each curve $X_i$ is observed with measurements errors on a grid $T_i=(t_{i1},\ldots,t_{iK})$, i.e. we observe a sample of $K$-dimensional vectors $Y_1,\ldots,Y_n$ with $Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}$, where the random variables $\epsilon_{ij}$ have mean zero and are uncorrelated with each other. We assume that each observation grid $T_1,\ldots,T_n$ is dense.

We begin by converting the discrete functional observations into continuous ones. Let $\tilde X_1,\ldots,\tilde X_n$ denote the functional versions of the raw data by some smoothing method. Our proposed methodology for estimating the pairwise geodesic distances $\{ d_\M (X_i,X_j)\}_{i>j}$ is agnostic to the smoothing method employed. For example we may employ spline smoothing to recover the functional versions of the raw data, i.e.\
$$ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}$$
where $\lambda>0$ is a tuning parameter controlling the smoothness of $\tilde X_i$.

Before introducing our procedure, let us first review the ISOMAP algorithm [@Tenenbaum2000], a three-step procedure that takes a set of points $x_1,\ldots,x_n$ in $(\M,g)$, a submanifold of $\R^D$ as input and produces an embedding of the input data in the space $\R^d$ with $d<D$, that preserves pairwise geodesic distances.

\begin{enumerate}
\item The first step constructs a weighted graph $G$ with nodes corresponding to the input data and for which the weight of an edge between two nodes $x_i$ and $x_j$ is equal to $||x_i-x_j ||_{\R^D}$. The graph $G$ is constructed such that a node is connected to a fix number $N$ of neighbors or to all nodes that are at a distance smaller than a given value $\epsilon$.
\item In the second step, the pairwise geodesic distances $d_g(x_i,x_j)$ are estimated based on $G$ using shortest-path algorithms. Specifically, the geodesic distance between two nodes is estimated to be the length of the shortest path in $G$ between these two nodes, which we may calculate for instance with the Floyd-Warshall algorithm [@Floyd1962].
<!-- Which one do we actually use in our implementation, Floyd-Warshall or Dijkstra's? -->
\item The last step consists in using multidimensional scaling to obtain an embedding in $\R^d$ that preserves the pairwise geodesic distances estimated above.
\end{enumerate}

In what follows, we call IsoGeo the procedure which performs the first two steps of ISOMAP.Note that shortest-path algorithms such as the Floyd-Warshall algorithm are used to find the smallest path given a weigthed graph but they do not produce the weighted graphs themselves. In fact, P-ISOMAP and robust ISOMAP are both modifications of ISOMAP in the sense that they change the construction of the weighted graph in Step 1.  
<!-- (which is the difficult part, depend on number of neighbors, many method out there to calculate that graph in a robust way, P-ISOMAP is one of these). -->

Our method is based on the idea that the underlying functional manifold can be sufficiently well-recovered by the subspace-constrained mean-shift algorithm [@Ozertem2011]. Theoretical justificaton for this can be found in [@Genovese2014].  and then performing operations on this hidden manifold. We expect the points $\tilde X^{s,\hat \M}_1,\ldots,\tilde X^{s,\hat \M}_n$ to lie close to the real manifold $\M$ and then $d_{\hat \M}$ to be close to $d_\M$. is described by the following steps:

<!-- 1. [Susan: I think the smoothing step should be independent of our proposed methodology] Transform each vector $Y_i$ into a function $\tilde X_i$ by spline smoothing: -->
<!-- $$ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}$$ -->
<!-- where $\lambda>0$ is a tuning parameter controlling the smoothness of $\tilde X_i$.  -->
\begin{enumerate}
\item Discretise the functions $\tilde X_1,\ldots,\tilde X_n$ by evaluating them on a common grid $\tilde T=(t_{1},\ldots,t_{K})$. Let $s$ be a positive integer, smaller than $K$. Obtain $\tilde X^s_1,\ldots,\tilde X^s_n \in \mathbb R^s$ using multidimensional scaling whereby the pairwise $\mathbb R^K$ Euclidean distances on the discretised functions $\tilde X_i$ are preserved.
\item Apply the subspace constrained mean-shift algorithm [@Ozertem2011] to each of $\tilde X^s_i$ to obtain  $\tilde X^{s,\hat \M}_i$.
\item Use IsoGeo to approximate the pairwise geodesic distances $\{d_{\hat \M}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j)\}_{i>j}$.
\end{enumerate}
Thus our geodesic distance estimators is the $n{\times}n$ matrix $\hat G$ whose elements are given by
$$
\hat G(i,j)=\hat G(j,i) = \left\{ \begin{array}{ll}
 d_{\hat \M}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j) & \textrm{if $i\neq j$,}\\
 0 & \textrm{otherwise.}
  \end{array} \right.
$$
Since ridge estimation suffers from the curse of dimensionality, we first reduce the dimension of our data with multidimensional scaling in Step 1 before applying the mean-shift algorithm.

## Selection of tuning parameters
Describe heuristics. We pick the bandwidth $h$ in the subspace constrained mean-shift using Equation (A1) of [@ChenHo2015].

@@

* the near isometry metric: this is the area under the receiver operating curve with $\epsilon$ on the $x$-axis and the degree to which near-$\epsilon$ isometry holds on the $y$-axis. Near-$\epsilon$ isometry is a relaxation of isometry measured by the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.  
* the relative Frobenius metric: this is given by
$||d-\hat d||/||d||$
where $d$ and $\hat d$ are the true and estimated $n{\times}n$ geodesic distance matrices, respectively. 
* the Pearson correlation metric: this is the Pearson correlation coefficient between the upper diagonal of $d$ and $\hat d$.
